---
title: "GSE279086 Single-Cell Analysis - Part 4: Harmony Integration"
author: "Smriti Arora"
date: "`r Sys.Date()`"
output: html_document
---

# SCRIPT 04: HARMONY INTEGRATION AND FINAL PROCESSING

```{r setup}
library(Seurat)
library(harmony)
library(ggplot2)
library(patchwork)
library(SingleCellExperiment)
library(zellkonverter)
library(FNN)

# Set directories
outputDir <- "~/GSE279086/output/"
plotDir <- "~/GSE279086/plots/"
```

## STEP 1: Load Pre-Harmony Object and Join Layers

```{r load_and_join}
cat("\n", rep("=", 60), "\n", sep = "")
cat("LOADING PRE-HARMONY OBJECT\n")
cat(rep("=", 60), "\n", sep = "")

seurat_processed <- readRDS(file.path(outputDir, "03_GSE279086_seurat_pca_umap.rds"))

cat("Loaded object:\n")
print(seurat_processed)
cat("\nLayers present:", Layers(seurat_processed), "\n")

# Join layers if needed
if(length(Layers(seurat_processed)) > 3) {
  cat("\nJoining layers...\n")
  seurat_processed <- JoinLayers(seurat_processed)
  cat("After joining - Layers:", Layers(seurat_processed), "\n")
  
  # Save joined version
  saveRDS(seurat_processed, 
          file.path(outputDir, "03_GSE279086_seurat_pca_umap_JOINED.rds"))
  cat("✓ Saved joined version\n")
}
```

## STEP 2: Run Harmony Integration

```{r run_harmony}
cat("\n", rep("=", 60), "\n", sep = "")
cat("RUNNING HARMONY INTEGRATION\n")
cat(rep("=", 60), "\n", sep = "")

# Run Harmony batch correction
harmony_processed <- RunHarmony(
  seurat_processed,
  group.by.vars = "orig.ident",  # Batch correct by sample
  plot_convergence = TRUE,
  reduction = "pca",
  dims.use = 1:50,
  max.iter.harmony = 20,
  verbose = TRUE
)

cat("✓ Harmony integration complete\n")
```

## STEP 3: Run UMAP on Harmony-Corrected Data

```{r harmony_umap}
cat("\n", rep("=", 60), "\n", sep = "")
cat("RUNNING UMAP ON HARMONY REDUCTION\n")
cat(rep("=", 60), "\n", sep = "")

harmony_processed <- RunUMAP(
  harmony_processed, 
  reduction = "harmony", 
  dims = 1:50,
  reduction.name = "umap",
  reduction.key = "UMAP_"
)

cat("✓ UMAP complete\n")
```

## STEP 4: Clustering on Harmony-Integrated Data

```{r harmony_clustering}
cat("\n", rep("=", 60), "\n", sep = "")
cat("CLUSTERING ON HARMONY DATA\n")
cat(rep("=", 60), "\n", sep = "")

# Find neighbors
harmony_processed <- FindNeighbors(
  harmony_processed, 
  reduction = "harmony", 
  dims = 1:50
)

# Find clusters
harmony_processed <- FindClusters(
  harmony_processed, 
  resolution = 0.8
)

# Rename to Harmony_clusters
harmony_processed$Harmony_clusters <- harmony_processed$seurat_clusters

cat("✓ Clustering complete\n")
cat("Number of Harmony clusters:", length(unique(harmony_processed$Harmony_clusters)), "\n")
```

## STEP 5: Save Harmony-Integrated Object

```{r save_harmony}
cat("\n", rep("=", 60), "\n", sep = "")
cat("SAVING HARMONY-INTEGRATED OBJECT\n")
cat(rep("=", 60), "\n", sep = "")

saveRDS(harmony_processed, file.path(outputDir, "04_GSE279086_harmony_corrected.rds"))
cat("✓ Saved: 04_GSE279086_harmony_corrected.rds\n")
```

## STEP 6: Visualize Before vs. After Harmony

```{r visualize_harmony}
cat("\n", rep("=", 60), "\n", sep = "")
cat("VISUALIZING HARMONY INTEGRATION RESULTS\n")
cat(rep("=", 60), "\n", sep = "")

# Before Harmony - By Sample
p1 <- DimPlot(seurat_processed, reduction = "umap", group.by = "orig.ident") + 
  ggtitle("Before Harmony - By Sample") + 
  NoLegend()

# After Harmony - By Sample
p2 <- DimPlot(harmony_processed, reduction = "umap", group.by = "orig.ident") + 
  ggtitle("After Harmony - By Sample") + 
  NoLegend()

# Before Harmony - By Condition
p3 <- DimPlot(seurat_processed, reduction = "umap", group.by = "Condition") +
  ggtitle("Before Harmony - By Condition")

# After Harmony - By Condition
p4 <- DimPlot(harmony_processed, reduction = "umap", group.by = "Condition") +
  ggtitle("After Harmony - By Condition")

# Save individual plots
ggsave(file.path(plotDir, "GSE279086_Uncorrected_sample.png"), 
       p1, width = 7, height = 6, dpi = 600)
ggsave(file.path(plotDir, "GSE279086_Uncorrected_condition.png"), 
       p2, width = 8, height = 6, dpi = 600)
ggsave(file.path(plotDir, "GSE279086_Harmony_sample.png"), 
       p3, width = 7, height = 6, dpi = 600)
ggsave(file.path(plotDir, "GSE279086_Harmony_condition.png"), 
       p4, width = 8, height = 6, dpi = 600)

# Combined comparison plot
combined <- (p1 | p2) / (p3 | p4)
ggsave(file.path(plotDir, "harmony_comparison.png"),
       combined, width = 14, height = 12, dpi = 300)

cat("✓ Saved comparison plots\n")
```

## STEP 7: Batch Mixing Analysis

```{r batch_mixing}
cat("\n", rep("=", 60), "\n", sep = "")
cat("ANALYZING BATCH MIXING\n")
cat(rep("=", 60), "\n", sep = "")

# Function to compute batch mixing
compute_knn_batch_mixing <- function(seu, batch_var, reduction = "pca", dims = 1:50, k = 20){
  emb <- Embeddings(seu, reduction = reduction)[, dims, drop = FALSE]
  nn <- FNN::get.knn(emb, k = k)$nn.index
  labs <- seu@meta.data[[batch_var]]
  same <- sapply(seq_len(nrow(nn)), function(i) mean(labs[nn[i,]] == labs[i], na.rm = TRUE))
  df <- data.frame(batch = labs, frac_same_batch = same)
  aggregate(frac_same_batch ~ batch, df, mean)
}

batch_column <- "orig.ident"

mix_df <- compute_knn_batch_mixing(seurat_processed, batch_column, "pca", 1:50, 20)
mix_df_harmony <- compute_knn_batch_mixing(harmony_processed, batch_column, "harmony", 1:50, 20)

mix_df_combined <- rbind(
  transform(mix_df, Status = "Before"),
  transform(mix_df_harmony, Status = "Harmony")
)

batch_cor_plot <- ggplot(mix_df_combined, aes(x = batch, y = frac_same_batch, fill = Status)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Batch Mixing - GSE279086", x = "Samples", y = "Mean same-batch fraction") +
  scale_fill_manual(values = c("Before" = "#00CED1", "Harmony" = "#9370DB")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.title = element_blank())

ggsave(file.path(plotDir, "GSE279086_batch_mixing_barplot.png"), 
       batch_cor_plot, width = 18, height = 8, dpi = 300)

cat("✓ Saved batch mixing plot\n")
```

## STEP 8: Export to h5ad for CellTypist

```{r export_h5ad}
cat("\n", rep("=", 60), "\n", sep = "")
cat("EXPORTING TO H5AD FOR CELLTYPIST\n")
cat(rep("=", 60), "\n", sep = "")

# Join layers
cat("Joining layers...\n")
harmony_joined <- JoinLayers(harmony_processed)

# Remove scale.data to prevent memory issues
cat("Removing scale.data layer to prevent memory issues...\n")
harmony_joined[["RNA"]]@layers$scale.data <- NULL
cat("Remaining layers:", paste(Layers(harmony_joined), collapse = ", "), "\n")

# Convert to SingleCellExperiment
cat("Converting to SingleCellExperiment...\n")
sce_harmony <- as.SingleCellExperiment(harmony_joined, assay = "RNA")

# Write to h5ad with compression
cat("Writing h5ad file (this may take a few minutes)...\n")
writeH5AD(sce_harmony, 
          file = file.path(outputDir, "04_GSE279086_harmony_corrected.h5ad"),
          compression = "gzip")

# Display file size
h5ad_size <- file.size(file.path(outputDir, "04_GSE279086_harmony_corrected.h5ad")) / 1024^2
cat("✓ Saved: 04_GSE279086_harmony_corrected.h5ad\n")
cat("H5AD file size:", round(h5ad_size, 2), "MB\n")
```

## STEP 9: Export UMAP Coordinates

```{r export_umap}
# Export UMAP coordinates for downstream use
umap_coords <- Embeddings(harmony_processed, "umap")
write.csv(umap_coords, file.path(plotDir, "GSE279086_umap_coordinates.csv"))
cat("✓ Saved UMAP coordinates\n")
```

## STEP 10: Final Summary

```{r final_summary}
cat("\n", rep("=", 60), "\n", sep = "")
cat("✓ HARMONY INTEGRATION COMPLETE!\n")
cat(rep("=", 60), "\n", sep = "")

cat("\nFiles created:\n")
cat("1. 04_GSE279086_harmony_corrected.rds (for R)\n")
cat("2. 04_GSE279086_harmony_corrected.h5ad (for CellTypist/Python)\n")
cat("3. Harmony comparison plots\n")
cat("4. Batch mixing analysis plot\n")
cat("5. UMAP coordinates\n")

cat("\n=== FINAL OBJECT SUMMARY ===\n")
cat("Cells:", ncol(harmony_processed), "\n")
cat("Genes:", nrow(harmony_processed), "\n")
cat("Samples:", length(unique(harmony_processed$orig.ident)), "\n")
cat("Conditions: HC (", sum(harmony_processed$Condition == "HC"), 
    ") + T1D (", sum(harmony_processed$Condition == "T1D"), ")\n", sep = "")
cat("Harmony clusters:", length(unique(harmony_processed$Harmony_clusters)), "\n")

cat("\n🎉 Ready for CellTypist annotation!\n")
```

---

**Next Step:** Run CellTypist annotation in Python using the h5ad file
